---
title: "Graphics-Asha"
output: html_document
date: "2024-12-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("pander")
install.packages("lubridate")
install.packages("mosaic")
install.packages("ggplot2")

library(readr)
library(dplyr)
library(tidyverse)
library(lubridate)
library(mosaic)
library(ggplot2)
library(pander)

Data_73_21 <- read_csv("Data_73_21.csv", na=c("NR", "NA"), col_types = "ccnccccccn")
Data_22 <- read_csv("Data_22.csv", col_types = "ncccnccccnc", na=c("NR", "NA"))
Data_23 <- read_csv("Data_23.csv", col_types = "ncccnccccnc", na=c("NR", "NA"))
Data_24 <- read_csv("Data_24.csv", col_types = "ncccnccccnc", na=c("NR", "NA"))
```


Line graph showing average race times across different age groups over time. (Years 1973 to 2024)

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Create age groups (if not already grouped)
all_data <- all_data %>%
  mutate(Age_Group = case_when(
    Age < 20 ~ "<20",
    Age >= 20 & Age < 30 ~ "20-29",
    Age >= 30 & Age < 40 ~ "30-39",
    Age >= 40 & Age < 50 ~ "40-49",
    Age >= 50 ~ "50+"
  ))

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(!is.na(Time_in_Minutes) & !is.na(Age_Group) & !is.na(Year))

# Calculate average race time for each age group over years
average_race_times <- all_data %>%
  group_by(Year, Age_Group) %>%
  summarize(Average_Time = mean(Time_in_Minutes, na.rm = TRUE), .groups = "drop")

# Create the line graph
ggplot(data = average_race_times, aes(x = Year, y = Average_Time, color = Age_Group)) +
  geom_line(size = 1) +
  scale_x_continuous(breaks = seq(1973, 2024, 5), limits = c(1973, 2024)) +  # Set x-axis from 1973 to 2024
  labs(
    title = "Average Race Times Across Age Groups (1973–2024)",
    x = "Year",
    y = "Average Race Time (minutes)",
    color = "Age Group"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )



```

Faceted line plots for different genders to explore gender-specific trends. (Years 1973 to 2024)

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Create age groups (if not already grouped)
all_data <- all_data %>%
  mutate(Age_Group = case_when(
    Age < 20 ~ "<20",
    Age >= 20 & Age < 30 ~ "20-29",
    Age >= 30 & Age < 40 ~ "30-39",
    Age >= 40 & Age < 50 ~ "40-49",
    Age >= 50 ~ "50+"
  ))

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(
    !is.na(Time_in_Minutes) & 
    !is.na(Year) & 
    !is.na(Division) & 
    !is.na(Age_Group)
  )

# Calculate average race time for each age group, year, and division
average_race_times <- all_data %>%
  group_by(Year, Division, Age_Group) %>%
  summarize(Average_Time = mean(Time_in_Minutes, na.rm = TRUE), .groups = "drop")

# Create the faceted line graph
ggplot(data = average_race_times, aes(x = Year, y = Average_Time, color = Age_Group)) +
  geom_line(size = 1) +
  scale_x_continuous(breaks = seq(1973, 2024, 5), limits = c(1973, 2024)) +  # Set x-axis from 1973 to 2024
  labs(
    title = "Average Race Times Across Age Groups by Division (1973–2024)",
    x = "Year",
    y = "Average Race Time (minutes)",
    color = "Age Group"
  ) +
  facet_wrap(~Division) +  # Facet by Division (gender proxy)
  theme_bw() +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )

```

Spaghetti plot of individual runners’ pace over multiple years. Highlight notable repeat runners. (Years 1973 to 2024)
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(
    !is.na(Time_in_Minutes) & 
    !is.na(Year) & 
    !is.na(Name)
  )

# Identify repeat runners (participated in multiple years)
repeat_runners <- all_data %>%
  group_by(Name) %>%
  summarize(Years_Run = n_distinct(Year), .groups = "drop") %>%
  filter(Years_Run > 2) %>%  # Define "notable" as participating in more than 2 years
  pull(Name)

# Filter data for repeat runners
all_data <- all_data %>%
  mutate(Is_Notable = ifelse(Name %in% repeat_runners, "Notable Repeat Runner", "Other Runner"))

# Create the spaghetti plot
ggplot(all_data, aes(x = Year, y = Time_in_Minutes, group = Name)) +
  geom_line(aes(color = Is_Notable), alpha = 0.6, size = 0.8) +
  scale_color_manual(
    values = c("Notable Repeat Runner" = "red", "Other Runner" = "lightblue"),
    name = "Runner Type"
  ) +
  labs(
    title = "Spaghetti Plot of Individual Runners' Pace Over Multiple Years",
    x = "Year",
    y = "Pace (minutes per mile)"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )
```

Scatter plot showing change in performance over time for runners who participated multiple times. (Years 1973 to 2024)
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(
    !is.na(Time_in_Minutes) & 
    !is.na(Year) & 
    !is.na(Name)
  )

# Identify runners who participated multiple times
repeat_runners <- all_data %>%
  group_by(Name) %>%
  summarize(Years_Run = n_distinct(Year), .groups = "drop") %>%
  filter(Years_Run > 1) %>%  # Only include runners with more than 1 participation year
  pull(Name)

# Filter data for repeat runners
repeat_runners_data <- all_data %>%
  filter(Name %in% repeat_runners)

# Calculate year-over-year change in performance
repeat_runners_data <- repeat_runners_data %>%
  arrange(Name, Year) %>%  # Ensure data is ordered by runner and year
  group_by(Name) %>%
  mutate(Change_in_Performance = Time_in_Minutes - lag(Time_in_Minutes)) %>%
  ungroup()

# Create the scatter plot
ggplot(repeat_runners_data, aes(x = Year, y = Change_in_Performance, color = Name)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Change in Performance Over Time for Runners Who Participated Multiple Times",
    x = "Year",
    y = "Change in Performance (minutes)",
    color = "Runner"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 12),
    legend.position = "none"  # Hide legend for readability with many runners
  )
```

Scatter plot of age vs. average pace for all runners. (Years 1973 to 2024)
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(
    !is.na(Time_in_Minutes) & 
    !is.na(Age) & 
    !is.na(Year)
  )

# Calculate average pace for each runner
average_pace_data <- all_data %>%
  group_by(Name, Age) %>%
  summarize(Average_Pace = mean(Time_in_Minutes, na.rm = TRUE), .groups = "drop")

# Create the scatter plot
ggplot(average_pace_data, aes(x = Age, y = Average_Pace)) +
  geom_point(alpha = 0.6, color = "lightblue", size = 2) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Optional linear trend line
  labs(
    title = "Scatter Plot of Age vs. Average Pace for All Runners (1973–2024)",
    x = "Age (years)",
    y = "Average Pace (minutes per mile)"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 12)
  )
```

Box plot comparing pace distributions across age groups. (Years 1973 to 2024)
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Create age groups
all_data <- all_data %>%
  mutate(Age_Group = case_when(
    Age < 20 ~ "<20",
    Age >= 20 & Age < 30 ~ "20-29",
    Age >= 30 & Age < 40 ~ "30-39",
    Age >= 40 & Age < 50 ~ "40-49",
    Age >= 50 ~ "50+"
  ))

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(
    !is.na(Time_in_Minutes) & 
    !is.na(Age_Group) & 
    !is.na(Year)
  )

# Create the box plot
ggplot(all_data, aes(x = Age_Group, y = Time_in_Minutes)) +
  geom_boxplot(fill = "lightblue", outlier.color = "red", outlier.shape = 1) +
  labs(
    title = "Pace Distributions Across Age Groups (1973–2024)",
    x = "Age Group",
    y = "Pace (minutes per mile)"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 12)
  )
```

Histogram of finishing times for each year. (Years 1973 to 2024)
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(
    !is.na(Time_in_Minutes) & 
    !is.na(Year)
  )

# Create the histogram faceted by year
ggplot(all_data, aes(x = Time_in_Minutes)) +
  geom_histogram(binwidth = 10, fill = "light blue", color = "light blue", alpha = 0.7) +
  facet_wrap(~Year, scales = "free_y") +
  labs(
    title = "Histogram of Finishing Times by Year (1973–2024)",
    x = "Finishing Time (minutes)",
    y = "Count"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 12),
    strip.text = element_text(size = 10),
    panel.spacing = unit(0.5, "lines")
  )
```

Identify trends in runners completing just before the cutoff time. (Years 1973 to 2024)

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(
    !is.na(Time_in_Minutes) & 
    !is.na(Year)
  )

# Define cutoff as the 90th percentile of finishing times for each year
cutoff_data <- all_data %>%
  group_by(Year) %>%
  mutate(Cutoff_Time = quantile(Time_in_Minutes, 0.9, na.rm = TRUE)) %>%  # 90th percentile
  ungroup()

# Identify runners finishing close to the cutoff time
cutoff_data <- cutoff_data %>%
  mutate(Finishing_Close_To_Cutoff = Time_in_Minutes >= (Cutoff_Time - 10) & Time_in_Minutes <= Cutoff_Time)

# Calculate the proportion of runners finishing close to the cutoff for each year
cutoff_trends <- cutoff_data %>%
  group_by(Year) %>%
  summarize(
    Total_Runners = n(),
    Close_To_Cutoff = sum(Finishing_Close_To_Cutoff, na.rm = TRUE),
    Proportion_Close_To_Cutoff = Close_To_Cutoff / Total_Runners,
    .groups = "drop"
  )

# Visualize the trend over time
ggplot(cutoff_trends, aes(x = Year, y = Proportion_Close_To_Cutoff)) +
  geom_line(color = "lightblue", size = 1) +
  geom_point(size = 2, color = "red") +
  labs(
    title = "Trend of Runners Finishing Close to Cutoff Time (1973–2024)",
    x = "Year",
    y = "Proportion of Runners Close to Cutoff"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 12)
  )
```

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(
    !is.na(Time_in_Minutes) & 
    !is.na(Year)
  )

# Define cutoff as the 90th percentile of finishing times for each year
cutoff_data <- all_data %>%
  group_by(Year) %>%
  mutate(Cutoff_Time = quantile(Time_in_Minutes, 0.9, na.rm = TRUE)) %>%  # 90th percentile
  ungroup()

# Identify runners finishing close to the cutoff time
cutoff_data <- cutoff_data %>%
  mutate(Finishing_Close_To_Cutoff = Time_in_Minutes >= (Cutoff_Time - 10) & Time_in_Minutes <= Cutoff_Time)

# Calculate the number of runners finishing close to the cutoff for each year
cutoff_trends <- cutoff_data %>%
  group_by(Year) %>%
  summarize(
    Total_Runners = n(),
    Close_To_Cutoff = sum(Finishing_Close_To_Cutoff, na.rm = TRUE),  # Actual count
    .groups = "drop"
  )

# Visualize the trend over time
ggplot(cutoff_trends, aes(x = Year, y = Close_To_Cutoff)) +
  geom_line(color = "lightblue", size = 1) +
  geom_point(size = 2, color = "red") +
  labs(
    title = "Number of Runners Finishing Close to Cutoff Time (1973–2024)",
    x = "Year",
    y = "Number of Runners Close to Cutoff"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 12)
  )
```


Scatter plot showing potential non-finishers (e.g., finishing at the race’s cutoff time). (Years 1973 to 2024)
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Convert Time to numeric (in minutes)
all_data <- all_data %>%
  mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time))) / 60)

# Filter out rows with NA in critical columns
all_data <- all_data %>%
  filter(
    !is.na(Time_in_Minutes) & 
    !is.na(Year) &
    !is.na(Name)  # Assuming 'Name' is critical for identifying individuals
  )

# Define potential non-finishers as those finishing in the 95th percentile or later
potential_non_finishers <- all_data %>%
  group_by(Year) %>%
  mutate(Percentile95 = quantile(Time_in_Minutes, 0.95, na.rm = TRUE)) %>%
  filter(Time_in_Minutes >= Percentile95) %>%
  ungroup()

# Create the scatter plot
ggplot(potential_non_finishers, aes(x = Year, y = Time_in_Minutes, color = as.factor(Year))) +
  geom_point(alpha = 0.6, size = 2) +
  labs(
    title = "Finishers Outliers Over Time (1973–2024)",
    x = "Year",
    y = "Finishing Time (minutes)",
    color = "Year"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 12),
    legend.position = "right"
  )
```


Fastest runner ever by Division (Years 1973 to 2024)
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Combine all datasets into one
all_data <- bind_rows(Data_73_21, Data_22, Data_23, Data_24)

# Check and convert the 'Time' column to numeric (in minutes)
if("Time" %in% colnames(all_data)) {
    all_data <- all_data %>%
        mutate(Time_in_Minutes = as.numeric(period_to_seconds(hms(Time, quiet = TRUE))) / 60)
} else {
    stop("The 'Time' column does not exist in the dataset.")
}

# Filter out rows with NA in critical columns
all_data <- all_data %>%
    filter(
        !is.na(Time_in_Minutes) & 
        !is.na(Year) &
        !is.na(Division) &
        !is.na(Name)
    )

# Find the fastest runner ever in each division
fastest_runners_by_division <- all_data %>%
    group_by(Division) %>%
    summarize(Fastest_Time = min(Time_in_Minutes), .groups = "drop")

# Re-join to get details of the fastest runners
fastest_runners_by_division <- all_data %>%
    right_join(fastest_runners_by_division, by = c("Division", "Time_in_Minutes" = "Fastest_Time")) %>%
    distinct(Division, .keep_all = TRUE) %>%
    select(Division, Name, Year, Time_in_Minutes)

# Output the results
print(fastest_runners_by_division)
```


Box plots comparing the pace of repeat runners vs. single-race participants. (Years 1973 to 2024)
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Assuming all_data is already loaded and includes necessary columns: 'Name', 'Year', 'Time'
# Convert Time to numeric (in minutes)
all_data <- all_data %>%
    mutate(
        Time_in_Minutes = as.numeric(period_to_seconds(hms(Time, quiet = TRUE))) / 60,
        Distance = ifelse(Year == 2015, 9.1, 10),  # Conditional distance based on the year
        Pace_in_Minutes = Time_in_Minutes / Distance  # Calculate pace
    )

# Filter out rows with NA in critical columns
all_data <- all_data %>%
    filter(
        !is.na(Time_in_Minutes) & 
        !is.na(Year) &
        !is.na(Name) &
        !is.na(Pace_in_Minutes)
    )

# Example: Proceed with analysis or visualization
# For instance, you might want to visualize average pace trends for all participants
average_pace_per_year <- all_data %>%
    group_by(Year) %>%
    summarize(Average_Pace = mean(Pace_in_Minutes, na.rm = TRUE))

# Create a line graph to display the trends
ggplot(average_pace_per_year, aes(x = Year, y = Average_Pace)) +
    geom_line(color = "lightblue", size = 1) +
    geom_point(color = "red", size = 2) +
    labs(
        title = "Average Pace Trends (1973–2024)",
        x = "Year",
        y = "Average Pace (minutes per mile)"
    ) +
    theme_bw() +
    theme(
        text = element_text(size = 12)
    )
```

Sankey diagram (or equivalent) showing how runners transition between age groups over the years. (Years 1973 to 2024)
```{r}
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)
library(dplyr)
library(lubridate)

# Ensure the dataset is loaded and includes the 'Year' and 'Age' columns
# Here's a command to simulate loading data, replace with your actual data loading method
# all_data <- read.csv("path_to_your_data.csv")

# Calculate the average age per year
average_age_per_year <- all_data %>%
  filter(!is.na(Age) & !is.na(Year)) %>%  # Make sure to filter out missing values
  group_by(Year) %>%
  summarize(Average_Age = mean(Age, na.rm = TRUE)) %>%
  ungroup()

# Check the first few rows to ensure it looks correct
head(average_age_per_year)

# Generate a line graph of average age over the years
age_plot <- ggplot(average_age_per_year, aes(x = Year, y = Average_Age)) +
  geom_line(color = "lightblue", size = 1) +  # Draw lines
  geom_point(color = "red", size = 2) +  # Add points
  labs(title = "Average Age of Runners Over the Years",
       x = "Year",
       y = "Average Age") +
  theme_bw() +
  theme(text = element_text(size = 12))

# Print the plot
print(age_plot)
```

What percentage of runners run for more than 5 years straight  (Years 1973 to 2024)

```{r}

# Ensure data is sorted by Name and Year for consecutive checking
all_data <- all_data %>%
  arrange(Name, Year)
# Identify consecutive participation
all_data <- all_data %>%
  group_by(Name) %>%
  mutate(Consecutive = Year - lag(Year, default = Year[1] - 1) == 1) %>%
  ungroup()
all_data <- all_data %>%
  group_by(Name) %>%
  mutate(Consecutive_Streak = cumsum(!Consecutive)) %>%
  group_by(Name, Consecutive_Streak) %>%
  mutate(Year_Count = n()) %>%
  ungroup()
# Calculate runners with more than 5 consecutive years
runners_with_5plus_years <- all_data %>%
  group_by(Name) %>%
  summarize(Max_Streak = max(Year_Count)) %>%
  filter(Max_Streak > 5)

# Calculate percentage
total_runners <- n_distinct(all_data$Name)
percentage_with_5plus_years <- nrow(runners_with_5plus_years) / total_runners * 100

# Output the result
percentage_with_5plus_years
```






